---
title: "Introduction to CoRpower"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to CoRpower}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `CoRpower` package performs power calculations for correlates of risk, as described in [Gilbert, Janes, and Huang (2016). Power/Sample Size Calculations for Assessing Correlates of Risk in Clinical Efficacy Trials] The power calculations accomodate three types of biomarkers: 

- trichotomous
- dichotomous
- continuous,

as well as two types of sampling design:

- without replacement sampling for retrospective case-control designs
- Bernoulli sampling for prospective case-cohort designs.
  
The vignette aims to illustrate distinct features of the functions in the package (with some mathematical background) by walking through a number of power calculation scenarios for different biomarker types, sampling designs, and input parameters.

The functions included in this package are:

- `computeN()`
- `computePower()`
- `plotPowerTri()`
- `plotPowerCont()`
- `plotRRgradVE()`
- `plotVElatCont()`

## Set-up and notation | Without replacement sampling 

- Assume a randomized vaccine vs. placebo/control vaccine efficacy trial
- $N_1=$ number of vaccine recipients observed to be at risk at $\tau$
- $n_{cases,1}$ ($n_{controls,1}$) $=$ number of observed cases (controls) in vaccine recipients between $\tau$ and $\tau_{\mathrm{max}}$, where cases have $\Delta Y = 1$ and controls have $\Delta (1-Y) = 1$
    - $n_{cases,1} + n_{controls,1} \leq N_1$
- $n^S_{cases,1}$ ($n^S_{controls,1}$) $=$ number of observed cases (controls) in vaccine recipients between $\tau$ and $\tau_{\mathrm{max}}$ with measured $S(1)$ (or $S^{\ast}(1)$)

- If calculations done at design stage, then $N_1$, $n_{cases,1}$, $n_{controls,1}$, $n^S_{cases,1}$, and $n^S_{controls,1}$ projected numbers

## Algorithm for trichotomous biomarker $S(1)$ | Without replacement sampling 

1. Specify overall $VE$ between $\tau$ and $\tau_{\mathrm{max}}$
    + Protocol-specified design alternative or $\widehat{VE}$
2. Specify $risk_0$
    + Protocol-specified placebo-group endpoint rate or $\widehat{risk}_0$
3. Select a grid of $VE^{lat}_0$ values
    + E.g., ranging from $VE$ ($H_0$) to 0 (maximal $H_1$ not allowing harm by vaccination)
4. Select a grid of $VE^{lat}_1 \geq VE^{lat}_0$ values
    + E.g., $VE^{lat}_1 = VE$
5. Specify $P^{lat}_0$ and $P^{lat}_2$, then $P^{lat}_1 = 1-P^{lat}_0-P^{lat}_2$
    + Assuming $risk^{lat}_0(x) = risk_0$,
      $$VE = VE^{lat}_0 P^{lat}_0 + VE^{lat}_1 P^{lat}_1 + VE^{lat}_2 P^{lat}_2$$
      yields $VE^{lat}_2$
    + If $VE^{lat}_0$ varies from $VE$ to 0, then $VE^{lat}_2$ varies from VE to $VE\, (P^{lat}_0 + P^{lat}_2)/P^{lat}_2$
6. Specify $P_0$ and $P_2$, then $P_1 = 1-P_0-P_2$
    + E.g., $P_0 = P^{lat}_0$ and $P_2 = P^{lat}_2$
7. **Approach 1:** Specify two of the three $(Sens, FP^0, FP^1)$ and two of the three $(Spec, FN^2, FN^1)$
    + E.g., specify $Sens$ and $Spec$ and $FP^0 = FN^2 = 0$
    
    **Approach 2:** Specify $\sigma^2_{obs}$ and $\rho = \sigma^2_{tr} / \sigma^2_{obs}$ and determine $(Sens, Spec, FP^0, FP^1, FN^1, FN^2)$ (see below)
    + Typically, $\sigma^2_{obs} = 1$
    + **Calculation of $(Sens, Spec, FP^0, FP^1, FN^1, FN^2)$**
        <ol type="i">
        <li> Assuming the classical measurement error model, where $X^{\ast} \sim \mathsf{N}(0,\rho\sigma^2_{obs})$, solve
        $$P^{lat}_0 = P(X^{\ast} \leq \theta_0) \quad \textrm{and} \quad P^{lat}_2 = P(X^{\ast} > \theta_2)$$
        for $\theta_0$ and $\theta_2$
        <li> Generate $B$ realizations of $X^{\ast}$ and $S^{\ast} = X^{\ast} + e$, where $e \sim \mathsf{N}(0,(1-\rho)\sigma^2_{obs})$, and
        $X^{\ast}$ independent of $e$
            + $B = 20,000$ by default
        <li> Using $\theta_0$ and $\theta_2$ from Step i., define
        \[
        \begin{align}
        Spec(\phi_0) &= P(S^{\ast} \leq \phi_0 \mid X^{\ast} \leq \theta_0)\\
        FN^1(\phi_0) &= P(S^{\ast} \leq \phi_0 \mid X^{\ast} \in (\theta_0,\theta_2])\\
        FN^2(\phi_0) &= P(S^{\ast} \leq \phi_0 \mid X^{\ast} > \theta_2)\\
        Sens(\phi_2) &= P(S^{\ast} > \phi_2 \mid X^{\ast} > \theta_2)\\
        FP^1(\phi_2) &= P(S^{\ast} > \phi_2 \mid X^{\ast} \in (\theta_0,\theta_2])\\
        FP^0(\phi_2) &= P(S^{\ast} > \phi_2 \mid X^{\ast} \leq \theta_0)
        \end{align}
        \]
              
        Estimate $Spec(\phi_0)$ by
        $$\widehat{Spec}(\phi_0) = \frac{\#\{S^{\ast}_b \leq \phi_0, X^{\ast}_b \leq \theta_0\}}{\#\{X^{\ast}_b \leq \theta_0\}}\,$$ etc.
        <li> Find $\phi_0 = \phi^{\ast}_0$ and $\phi_2 = \phi^{\ast}_2$ that numerically solve
        \[
        \begin{align}
        P_0 &= \widehat{Spec}(\phi_0)P^{lat}_0 + \widehat{FN}^1(\phi_0)P^{lat}_1 + \widehat{FN}^2(\phi_0)P^{lat}_2\\
        P_2 &= \widehat{Sens}(\phi_2)P^{lat}_2 + \widehat{FP}^1(\phi_2)P^{lat}_1 + \widehat{FP}^0(\phi_2)P^{lat}_0
        \end{align}
        \]
        and compute
        \[
        Spec = \widehat{Spec}(\phi^{\ast}_0),\; Sens = \widehat{Sens}(\phi^{\ast}_2),\; \textrm{etc.}
        \]
        </ol>    
    + In Approach 2, plot
            \[
            RR_t \quad \textrm{vs.} \quad \frac{RR^{lat}_2}{RR^{lat}_0},
            \]
      where
      \[
      RR_t := \frac{risk_1(2)}{risk_1(0)} = \frac{\sum_{x=0}^2 RR^{lat}_x P(X=x|S(1)=2)}{\sum_{x=0}^2 RR^{lat}_x P(X=x|S(1)=0)}
      \]
    + For $\rho < 1$, $RR_t$ closer to 1 than $RR^{lat}_2\big/ RR^{lat}_0$

8. Simulate $M$ data sets under the true parameter values:
    <ol type="a">
    *Full data*
    <li> $N_x = (n_{controls,1} + n_{cases,1}) P^{lat}_x$
    <li> $\left(n_{cases,1}(0),n_{cases,1}(1),n_{cases,1}(2)\right) \sim \mathsf{Mult}(n_{cases,1},(p_0,p_1,p_2))$, where $p_k=P(X=k|Y=1,Y^{\tau}=0,Z=1)$
    <li> For each of the $N_x$ subjects, generate $S_i(1)$ by a draw from $\mathsf{Mult}(1,(p_0,p_1,p_2))$, where $p_k=P(S(1)=k|X=x)$ </li>
      
    *Observed data*
    <li> Sample without replacement $n^S_{cases,1}$ and $n^S_{controls,1} = K n^S_{cases,1}$ controls with measured $S(1)$ $(R=1)$, i.e., the control:case ratio is not fixed within subgroup $X=x$
    </ol>
9. For each observed data set, compute the 1-sided one-degree-of-freedom Wald test statistic for $H_0 \Leftrightarrow \{\widetilde{H}_0: \beta_S(1) \geq 0\}$ from IPW logistic regression model that accounts for biomarker sampling design (function `tps` in R package `osDesign`)
    + Alternatively, use the generalized two-degree-of-freedom Wald test
10. Compute power as proportion of data sets with 1-sided Wald test $p \leq \alpha/2$ for specified $\alpha$
11. Repeat power calculation varying control:case ratio, $n_{cases,1}$, $n^S_{cases,1}$, $(P^{lat}_0, P^{lat}_2, P_0, P_2)$, $(Sens, Spec)$, $\rho$
  
  
## Illustration: hypothetical randomized placebo-controlled VE trial
### **Trial design**
- $N_{rand} = 4,100$ participants randomized to each of the vaccine and placebo group and followed for $\tau_{\mathrm{max}}=$ 24 months
- Samples for measurement of $S(1)$ at $\tau=$ 3.5 months stored in all vaccine recipients
- Cumulative endpoint rate between $\tau$ and $\tau_{\mathrm{max}}$ in placebo group $=3.4\%$ $(=risk_0)$
- $VE_{\tau-\tau_{\mathrm{max}}}=75\%$, $\quad VE_{0-\tau}=VE_{\tau-\tau_{\mathrm{max}}}/2$
- Cumulative dropout rate between 0 and $\tau_{\mathrm{max}}$ in both groups $=10\%$

## Illustration: calculation of input parameters with `computeN()`
### Assumptions
<br><br>
<ol type="i">
<li> Failure time $T$ and censoring time $C$ are independent
<li> $T\mid Z=0 \sim \mathsf{Exp}(\lambda_T)$ and $C\mid Z=0 \sim \mathsf{Exp}(\lambda_C)$
<li> $RR_{\tau-\tau_{\mathrm{max}}} := \frac{P(T \leq \tau_{\mathrm{max}} \mid T>\tau, Z=1)}{P(T \leq \tau_{\mathrm{max}} \mid T>\tau, Z=0)} = \frac{P(T \leq t\mid T>\tau, Z=1)}{P(T \leq t\mid T>\tau, Z=0)}$ for all $t \in (\tau,\tau_{\mathrm{max}}]$
</ol>

### Number of vaccine recipients observed to be at risk at $\tau$
\[
\begin{align}
N_1 &= N_{rand}\, P(T>\tau, C>\tau \mid Z=1)\\
&= N_{rand}\, P(T>\tau \mid Z=1)\, P(C>\tau \mid Z=1)\\
&= N_{rand}\, \{1 - RR_{0-\tau}\, P(T \leq \tau \mid Z=0)\}\, P(C>\tau \mid Z=1)\\
&\approx 4,023
\end{align}
\]

### Number of observed cases in vaccine recipients between $\tau$ and $\tau_{\mathrm{max}}$
\[
\begin{align}
n_{cases,1} &= N_1\, P(T\leq \tau_{\mathrm{max}}, T\leq C \mid T>\tau, C>\tau, Z=1)\\
&= N_1\, P(T\leq \min(C,\tau_{\mathrm{max}}) \mid T>\tau, C>\tau, Z=1)\\
&= N_1\, \frac{\int_{\tau}^{\infty}P(\tau < T \leq \min(c,\tau_{\mathrm{max}})\mid Z=1)f_C(c)\mathop{\mathrm{d}}\!c}{P(T>\tau, C>\tau \mid Z=1)}\\
&= N_1\, \frac{\bigg\{\int_{\tau}^{\tau_{\mathrm{max}}}P(\tau < T \leq c\mid Z=1)f_C(c)\mathop{\mathrm{d}}\!c + P(\tau < T \leq \tau_{\mathrm{max}}\mid Z=1) P(C>\tau_{\mathrm{max}})\bigg\}}{P(T>\tau, C>\tau \mid Z=1)}\\
&\approx 32
\end{align}
\]

### Number of observed controls in vaccine recipients between $\tau$ and $\tau_{\mathrm{max}}$
\[
\begin{align}
n_{controls,1} &= N_1 \, P(T > \tau_{\mathrm{max}}, C > \tau_{\mathrm{max}} \mid T>\tau, C>\tau, Z=1)\\
&\approx 3,654
\end{align}
\]

### Number of observed cases (controls) in vaccine recipients between $\tau$ and $\tau_{\mathrm{max}}$ with measured $S(1)$
\[
\begin{align}
n^S_{cases,1} &= n_{cases,1}\\
n^S_{controls,1} &= K n^S_{cases,1}
\end{align}
\]

### Compute $N_1, n_{cases,1}, n_{controls,1},$ and $n^S_{cases,1}$ with `computeN()`
```{r}
library(CoRpower)
computeN(Nrand = 4100,          # participants randomized to vaccine arm
         tau = 3.5,             # biomarker sampling timepoint
         taumax = 24,           # end of follow-up
         VEtauToTaumax = 0.75,  # VE between 'tau' and 'taumax'
         VE0toTau = 0.75/2,     # VE between 0 and 'tau'
         risk0 = 0.034,         # placebo-group endpoint risk between 'tau' and 'taumax'
         dropoutRisk = 0.1,     # dropout risk between 0 and 'taumax'
         propCasesWithS = 1)    # proportion of observed cases with measured S(1)
```

## Illustration: `CoRpower` for trichotomous \(\, S(1)\) | Without replacement sampling
**Approach 1** $(Sens, Spec, FP^0, FN^2$ specified$)$

- **Scenario 1:** vary control:case ratio
- **Scenario 2:** vary $Sens$, $Spec$
- **Scenario 3:** vary $P_0^{lat}$, $P_2^{lat}$, $P_0$, $P_2$

**Approach 2** $(\sigma^2_{obs}$ and $\rho$ specified$)$

- **Scenario 4:** vary $\rho$
- **Scenario 5:** vary $P_0^{lat}$, $P_2^{lat}$, $P_0$, $P_2$
- **Scenario 6:** vary $n_{cases,1}$

## *Scenario 1:* vary control:case ratio (Approach 1) | Trichotomous $S(1)$, without replacement sampling
### Run simulations and compute power with `computePower()`
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,                 
                    nControlsTx = 3654,            
                    nCasesTxWithS = 32,           
                    controlCaseRatio = c(5, 3, 1), # n^S_controls : n^S_cases ratio         
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk from tau - taumax
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = 0.2,                  # prevalence of lower protected
                    Plat2 = 0.6,                  # prevalence of higher protected
                    P0 = 0.2,                     # probability of low biomarker response
                    P2 = 0.6,                     # probability of high biomarker response
                    sens = 0.8, spec = 0.8, FP0 = 0, FN2 = 0,
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "trichotomous")    # "continuous" by default
```

### Plot power curves with `plotPowerTri()`
Basic plotting functions are included in the package to aid with visualizing results. `plotPowerTri` plots the power curve against the CoR relative risk, $RR_t$, for trichotomous or binary biomarkers. 

Output from `computePower()` can be saved as an object and assigned to the `outComputePower` input parameter.  
```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr,  # 'computePower' output list of lists
             legendText = paste0("Control:Case = ", c("5:1", "3:1", "1:1")))
```

Alternatively, output from `computePower()` can be saved in RData files. In this case, the `outComputePower` input parameter should be the name(s) of the output file(s), and the `outDir` input parameter should be the name(s) of the file location(s). For more information, visit the `plotPowerTri()` help page. 
```{r, eval=FALSE}
computePower(..., saveDir = "myDir", saveFile = c("myFile1.RData", "myFile2.RData", "myFile3.RData"))
plotPowerTri(outComputePower = c("myFile1.RData", "myFile2.RData", "myFile3.RData"), # 'computePower' output files
             outDir = rep("~/myDir", 3),                           # path to each myFilex.RData
             legendText = paste0("Control:Case = ", c("5:1", "3:1", "1:1")))
```


## *Scenario 2:* vary \(\, Sens\) and \(\, Spec\) (Approach 1) | Trichotomous \(\, S(1)\), without replacement sampling
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,                 
                    nControlsTx = 3654,            
                    nCasesTxWithS = 32,           
                    controlCaseRatio = 5,         # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk from tau - taumax
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = 0.2,                  # prevalence of lower protected
                    Plat2 = 0.6,                  # prevalence of higher protected
                    P0 = 0.2,                     # probability of low biomarker response
                    P2 = 0.6,                     # probability of high biomarker response
                    sens = c(1, 0.9, 0.8, 0.7), spec = c(1, 0.9, 0.8, 0.7), 
                    FP0 = c(0, 0, 0, 0), FN2 = c(0, 0, 0, 0),
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "trichotomous")    # "continuous" by default
```

```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr,    
             legendText = paste0("Sens = Spec = ", c(1, 0.9, 0.8, 0.7)))
```

## *Scenario 3:* vary \(\, P^{lat}_0, P^{lat}_2, P_0, P_2\) (Approach 1) | Trichotomous \(\, S(1)\), without replacement sampling 
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,                 
                    nControlsTx = 3654,            
                    nCasesTxWithS = 32,           
                    controlCaseRatio = 5,         # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk from tau - taumax
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = c(0.05, 0.1, 0.15, 0.2),          
                    Plat2 = c(0.15, 0.3, 0.45, 0.6),          
                    P0 = c(0.05, 0.1, 0.15, 0.2),            
                    P2 = c(0.15, 0.3, 0.45, 0.6),          
                    sens = 0.8, spec = 0.8, FP0 = 0, FN2 = 0,
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "trichotomous")    # "continuous" by default
```

```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr, 
             legendText = c("Plat0=0.05, Plat2=0.15", 
                            "Plat0=0.1, Plat2=0.3", 
                            "Plat0=0.15, Plat2=0.45", 
                            "Plat0=0.2, Plat2=0.6"))
```

## *Scenario 4:* vary \(\, \rho \) (Approach 2) | Trichotomous \(\, S(1)\), without replacement sampling 
### Run simulations and compute power with `computePower()`
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,                 
                    nControlsTx = 3654,            
                    nCasesTxWithS = 32,           
                    controlCaseRatio = 5,         # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk from tau - taumax 
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = 0.2,                  # prevalence of lower protected
                    Plat2 = 0.6,                  # prevalence of higher protected
                    P0 = 0.2,                     # probability of low biomarker response
                    P2 = 0.6,                     # probability of high biomarker response
                    sigma2obs = 1,                # variance of observed biomarker S(1)
                    rho = c(1, 0.9, 0.7, 0.5),    # protection-relevant fraction of variance of S(1)
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "trichotomous")    # "continuous" by default
```

### Plot power curves with `plotPowerTri()`
```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr,   
             legendText = paste0("rho = ", c(1, 0.9, 0.7, 0.5)))
```

### Plot $RR_t$ vs. $RR_2^{lat}/RR_0^{lat}$ with `plotRRgradVE()`
`plotRRgradVE()` plots the ratio of relative risks for the higher and lower latent subgroups $RR_2^{lat}/RR_0^{lat}$ against the CoR relative risk effect size $RR_t = risk_1(2)/risk_1(0)$. 

Output from `computePower()` can be saved as an object and assigned to the `outComputePower` input parameter.
```{r, eval=FALSE}
plotRRgradVE(outComputePower = pwr,  # 'computePower' output list of lists
             legendText = paste0("rho = ", c(1, 0.9, 0.7, 0.5)))
```

Alternatively, output from `computePower()` can be saved in RData files. In this case, the `outComputePower` input parameter should be the name(s) of the output file(s), and the `outDir` input parameter should be the name(s) of the file location(s). For more information, visit the `plotRRgradVE()` help page. 
```{r, eval=FALSE}
computePower(..., saveDir = "myDir", saveFile = "myFile.RData")
plotRRgradVE(outComputePower = paste0("myFile_rho_", c(1, 0.9, 0.7, 0.5), ".RData"),    # files with 'computePower' output
             outDir = "~/myDir",            # path to myFile.RData
             legendText = paste0("rho = ", c(1, 0.9, 0.7, 0.5)))
```

### Plot ROC curves with `plotROCcurveTri()`
`plotROCcurveTri()` plots the receiver operating characteristic (ROC) curve displaying sensitivity and specificity for a range of values for `P2`, `P0`, `Plat2`, and `rho`. 
For more information, visit the `plotROCcurveTri()` help page. 
```{r, eval=FALSE}
plotROCcurveTri(Plat0 = 0.2, 
                Plat2 = c(0.2, 0.3, 0.4, 0.5), 
                P0 = seq(0.90, 0.10, len=25), 
                P2 = seq(0.10, 0.90, len=25), 
                rho = c(1, 0.9, 0.7, 0.5))
```

## *Scenario 5:* vary \(\, P^{lat}_0, P_0, P^{lat}_2, P_2 \) (Approach 2) | Trichotomous \(\, S(1)\), without replacement sampling
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,                 
                    nControlsTx = 3654,            
                    nCasesTxWithS = 32,           
                    controlCaseRatio = 5,         # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk from tau - taumax 
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = c(0.05, 0.1, 0.15, 0.2),          
                    Plat2 = c(0.15, 0.3, 0.45, 0.6),          
                    P0 = c(0.05, 0.1, 0.15, 0.2),            
                    P2 = c(0.15, 0.3, 0.45, 0.6), 
                    sigma2obs = 1,                # variance of observed biomarker S(1)
                    rho = 0.9,                    # protection-relevant fraction of variance of S(1)
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "trichotomous")    # "continuous" by default
```

```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr,  
             legendText = c("Plat0=0.05, Plat2=0.15", 
                            "Plat0=0.1, Plat2=0.3", 
                            "Plat0=0.15, Plat2=0.45", 
                            "Plat0=0.2, Plat2=0.6"))
```

## *Scenario 6:* vary \(\, n_{cases,1}\) (Approach 2) | Trichotomous \(\, S(1)\), without replacement sampling
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = c(25, 32, 35, 40),             
                    nControlsTx = c(3661, 3654, 3651, 3646),  
                    nCasesTxWithS = c(25, 32, 35, 40),       
                    controlCaseRatio = 5,         # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk fom tau - taumax
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = 0.2,                  # prevalence of lower protected
                    Plat2 = 0.6,                  # prevalence of higher protected
                    P0 = 0.2,                     # probability of low biomarker response
                    P2 = 0.6,                     # probability of high biomarker response
                    sigma2obs = 1,                # variance of observed biomarker S(1)
                    rho = 0.9,                    # protection-relevant fraction of variance of S(1)
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "trichotomous")    # "continuous" by default
```

```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr,   
             legendText = paste0("nCasesTx = ", c(25, 32, 35, 40)))
```

## Illustration: `CoRpower` for binary \(\, S(1)\) | Without replacement sampling
Achieved by selecting $P_0^{lat}$, $P_2^{lat}$, $P_0$, $P_2$ such that
\[
\begin{align}
P_0^{lat} + P_2^{lat} &= 1\\
P_0 + P_2 &= 1
\end{align}
\]

**Approach 2** $(\sigma^2_{obs}$ and $\rho$ specified$)$

- **Scenario 7:** vary $n_{cases,1}$

## **Scenario 7:** vary \(\, n_{cases,1}\) (Approach 2) | Binary \(\, S(1)\), without replacement sampling 
### Run simulations and compute power with `computePower()`
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = c(25, 32, 35, 40),             
                    nControlsTx = c(3661, 3654, 3651, 3646),  
                    nCasesTxWithS = c(25, 32, 35, 40),       
                    controlCaseRatio = 5,         # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk from tau - taumax
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = 0.2,                  # prevalence of lower protected
                    Plat2 = 0.8,                  # prevalence of higher protected
                    P0 = 0.2,                     # probability of low biomarker response
                    P2 = 0.8,                     # probability of high biomarker response
                    sigma2obs = 1,                # variance of observed biomarker S(1)
                    rho = 0.9,                    # protection-relevant fraction of variance of S(1)
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "binary")          # "continuous" by default
```

### Plot power curves with `plotPowerTri()`
```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr,   
             legendText = paste0("nCasesTx = ", c(25, 32, 35, 40)))
```

## Algorithm for continuous biomarker \(\, S^{\ast}(1)\) | Without replacement sampling
<ol>
<li> Specify overall $VE$ between $\tau$ and $\tau_{\mathrm{max}}$
  <ul>
  <li> Protocol-specified design alternative or $\widehat{VE}$
  </ul>
</li>
<li> Specify $risk_0$
  <ul>
  <li> Protocol-specified placebo-group endpoint rate or $\widehat{risk}_0$
  </ul>
</li>
<li> Specify $P^{lat}_{lowestVE}$, $\rho$, and a grid of $VE_{lowest}$ values (e.g., ranging from $VE$ to 0)
  <ul>
  <li> Fixed $(VE, risk_0, P^{lat}_{lowestVE}, VE_{lowest}, \rho)$ and
  \[
  \begin{align}
  risk^{lat}_1(x^{\ast}) &= (1 - VE_{lowest}) risk_0,\quad x^{\ast} \leq \nu\\
  \mathop{\mathrm{logit}}\{risk^{lat}_1(x^{\ast})\} &= \alpha^{lat}+\beta^{lat}x^{\ast},\quad x^{\ast} \geq \nu\\
  VE &= 1 - \frac{\int risk^{lat}_1(x^{\ast})\phi(x^{\ast}/\sqrt{\rho} \sigma_{obs})\mathop{\mathrm{d}}\!x^{\ast}}{risk_0}
  \end{align}
  \]
  yield $\alpha^{lat}$ and $\beta^{lat}$
  <li> Plot $VE^{lat}_{x^{\ast}}$ vs. $x^{\ast}$ and calculate the pertaining CoR effect size $\exp(\beta^{lat})$ for each level of $VE_{lowest}$
  </ul>
</li>
<li> Simulate $M$ data sets under the true parameter values:
  <ol type="a">
  *Full data*
  <li> Sample $X^{\ast}$ for $n_{cases,1}$ cases from $f_{X^{\ast}}(x^{\ast}|Y=1,Y^{\tau}=0,Z=1)$ using Bayes rule
  <li> Sample $X^{\ast}$ for $n_{controls,1}$ controls from $f_{X^{\ast}}(x^{\ast}|Y=0,Y^{\tau}=0,Z=1)$ using Bayes rule
  <br>
  - How? Use a fine grid of $\widetilde{x}^{\ast}$ values and then 
  <br>
  $\;\;$`sample(`$\widetilde{x}^{\ast}$, `prob=`$f_{X^{\ast}}(\widetilde{x}^{\ast}|Y=\cdot,Y^{\tau}=0,Z=1)$, `replace=TRUE)`
  <li> Sample $S^{\ast}(1)$ following $S^{\ast}(1) = X^{\ast} + e$ </li>
  *Observed data*
  <li> Sample without replacement $n^S_{cases,1}$ and $n^S_{controls,1} = K n^S_{cases,1}$ controls with measured $S^{\ast}(1)$ $(R=1)$
  </ol>
</li>
<li> For each observed data set, compute the 1-sided one-degree-of-freedom Wald test statistic for $H_0 \Leftrightarrow \{\widetilde{H}_0: \beta_{S^{\ast}(1)} \geq 0\}$ from IPW logistic regression model that accounts for biomarker sampling design (function `tps` in R package `osDesign`)
<li> Compute power as proportion of data sets with 1-sided Wald test $p \leq \alpha/2$ for specified $\alpha$
<li> Repeat power calculation varying control:case ratio, $n_{cases,1}$, $n^S_{cases,1}$, $P^{lat}_{lowestVE}$, $\rho$
</ol>

## Illustration: `CoRpower` for continuous \(\, S^{\ast}(1)\) | Without replacement sampling
- **Scenario 8:** vary $\rho$
- **Scenario 9:** vary $P_{lowestVE}^{lat}$

## *Scenario 8:* vary \(\, \rho \) | Continuous \(\, S^{\ast}(1)\), without replacement sampling 
### Run simulations and compute power with `computePower()`
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,             
                    nControlsTx = 3654,         
                    nCasesTxWithS = 32,         
                    controlCaseRatio = 5,        # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,            # overall VE
                    risk0 = 0.034,               # placebo-group endpoint risk from tau - taumax
                    PlatVElowest = 0.2,          # prevalence of VE_lowest
                    VElowest = seq(0, VEoverall, len=100), # lowest VE for true biomarker X*<=nu
                    sigma2obs = 1,               # variance of observed biomarker S
                    rho = c(1, 0.9, 0.7, 0.5)    # protection-relevant fraction of variance of S
                    M = 1000,                    # number of simulated clinical trials
                    alpha = 0.05,                # two-sided Wald test Type 1 error rate
                    biomType = "continuous")     # "continuous" by default
```

### Plot power curves with `plotPowerCont()`
`plotPowerCont()` plots the power curve against the CoR relative risk, $RR_c$, for continuous biomarkers.  

Output from `computePower()` can be saved as an object and assigned to the `outComputePower` input parameter. In this scenario, since `computePower()` was run multiple times to vary the controls:cases ratio, these multiple output objects can be read into the function as a list. 
```{r, eval=FALSE}
plotPowerCont(outComputePower = pwr,          # output list of lists from 'computePower'
              legendText = paste0("rho = ", c(1, 0.9, 0.7, 0.5)))
```

Alternatively, output from `computePower()` can be saved in RData files. In this case, the `outComputePower` input parameter should be the name(s) of the output file(s), and the `outDir` input parameter should be the name(s) of the file location(s). For more information, visit the `plotPowerCont()` help page.
```{r, eval=FALSE}
computePower(..., saveDir = "myDir", saveFile = "myFile.RData")
plotPowerCont(outComputePower = paste0("myFile_rho_", c(1, 0.9, 0.7, 0.5), ".RData"),     # files with 'computePower' output
              outDir = "~/myDir",             # path to myFile.RData
              legendText = paste0("rho = ", c(1, 0.9, 0.7, 0.5)))
```

### Plot $VE^{lat}_{x^{\ast}}$ curves with `plotVElatCont()`
`plotVElatCont()` plots the vaccine efficacy (VE) curve for the true biomarker X*=x* for eight different values of the true CoR relative risk, \eqn{RR_c (\rho=1)}, in vaccine recipients and the lowest vaccine efficacy level for the true biomarker, \eqn{VE_lowest}.

`outComputePower` contains output from a single run of `computePower()` with no varying argument (i.e., no vectorized input parameters other than $VE^{lat}_0$, $VE^{lat}_1$, and \eqn{VE_lowest}). This output can be in the form of an assigned object, which is a list of lists of length $1$, or a character string specifying the file containing the output. Note that this is unlike `plotPowerTri()` and `plotPowerCont()`, which can take in output from `computePower()` in the form of a list of lists of length greater than $1$ or a character vector. For more information, visit the `plotVElatCont()` help page. 

Using the function when `computePower()` output is saved as list object `pwr`:
```{r, eval=FALSE}
plotVElatCont(outComputePower = pwr)  
```

Using the function when `computePower()` output is saved in a file with name "myFile" and location "~/myDir":
```{r, eval=FALSE}
computePower(..., saveDir = "myDir", saveFile = "myFile.RData")
plotVElatCont(outComputePower = "myFile.RData",   
              outDir = "~/myDir")          
```

## *Scenario 9:* vary \(\, P_{lowestVE}^{lat} \) | Continuous \(\, S^{\ast}(1)\), without replacement sampling
### Run simulations and compute power with `computePower()`
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,               
                    nControlsTx = 3654,         
                    nCasesTxWithS = 32,         
                    controlCaseRatio = 5,        # n^S_controls : n^S_cases ratio
                    VEoverall = 0.75,            # overall VE
                    risk0 = 0.034,               # placebo-group endpoint risk from tau - taumax
                    PlatVElowest = c(0.05, 0.1, 0.15, 0.2),         
                    VElowest = seq(0, VEoverall, len=100), # lowest VE for true biomarker X*<=nu
                    sigma2obs = 1,               # variance of observed biomarker S(1)
                    rho = 0.9                    # protection-relevant fraction of variance of S(1)
                    M = 1000,                    # number of simulated clinical trials
                    alpha = 0.05,                # two-sided Wald test Type 1 error rate
                    biomType = "continuous")     # "continuous" by default
```

### Plot power curves with `plotPowerCont()`
```{r, eval=FALSE}
plotPowerCont(outComputePower = pwr,          # output list of lists from 'computePower'
              legendText = paste0("PlatVElowest = ", c(0.05, 0.1, 0.15, 0.2)))
```

## Bernoulli / case-cohort sampling of \(\, S(1)\) (or \(\, S^{\ast}(1)\))
- Bernoulli sample at baseline with sampling probability $p$
- $S(1)$ (or $S^{\ast}(1)$) measured at $\tau$ in
    - a subset of the sample with $Y^{\tau}=0$, and
    - all cases with $Y^{\tau}=0$
- Implications:
    - $n_{cases,1} = n^S_{cases,1}$
    - design parameter $n^S_{controls,1}$ replaced by probability $p$ because $n^S_{controls,1}$ is a random variable in case-cohort designs

## Illustration: `CoRpower` for trichotomous \(\, S(1)\) and continuous \(\, S^{\ast}(1)\) | Bernoulli sampling
Trichotomous $S(1)$ (Approach 1)

- **Scenario 10:** vary $p$

Continuous $S^{\ast}(1)$

- **Scenario 11:** vary $p$

## *Scenario 10:* vary \(\, p \) (Approach 1) | Trichotomous \(\, S(1) \), Bernoulli sampling
### Run simulations and compute power with `computePower()`
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,             
                    nControlsTx = 3654,       
                    nCasesTxWithS = 32,       
                    cohort = TRUE,                # FALSE by default
                    p = c(0.01, 0.02, 0.03, 0.05),               
                    VEoverall = 0.75,             # overall VE
                    risk0 = 0.034,                # placebo-group endpoint risk from tau - taumax
                    VElat0 = seq(0, VEoverall, len=100), # grid of VE (V/P) among lower protected
                    VElat1 = rep(VEoverall, 100), # grid of VE (V/P) among medium protected
                    Plat0 = 0.2,                  # prevalence of lower protected
                    Plat2 = 0.6,                  # prevalence of higher protected
                    P0 = 0.2,                     # probability of low biomarker response
                    P2 = 0.6,                     # probability of high biomarker response
                    sens = 0.8, spec = 0.8, FP0 = 0, FN2 = 0,
                    M = 1000,                     # number of simulated clinical trials
                    alpha = 0.05,                 # two-sided Wald test Type 1 error rate
                    biomType = "trichotomous")    # "continuous" by default
```

### Plot power curves with `plotPowerTri()`
```{r, eval=FALSE}
plotPowerTri(outComputePower = pwr,  # 'computePower' output
             legendText = paste0("Cohort p = ", c(0.01, 0.02, 0.03, 0.05)))
```

## *Scenario 11:* vary \(\, p \) | Continuous \(\, S^{\ast}(1)\), Bernoulli sampling
### Run simulations and compute power with `computePower()`
```{r, eval=FALSE}
pwr <- computePower(nCasesTx = 32,             
                    nControlsTx = 3654,        
                    nCasesTxWithS = 32,       
                    cohort = TRUE,               # FALSE by default
                    p = c(0.01, 0.02, 0.03, 0.05),                  
                    VEoverall = 0.75,            # overall VE
                    risk0 = 0.034,               # placebo-group endpoint risk from tau - taumax
                    PlatVElowest = 0.2,          # prevalence of VE_lowest
                    VElowest = seq(0, VEoverall, len=100), # lowest VE for true biomarker X*<=nu
                    sigma2obs = 1,               # variance of observed biomarker S(1)
                    rho = 0.9                    # protection-relevant fraction of variance of S(1)
                    M = 1000,                    # number of simulated clinical trials
                    alpha = 0.05,                # two-sided Wald test Type 1 error rate
                    biomType = "continuous")     # "continuous" by default
```

### Plot power curves with `plotPowerCont()`
```{r, eval=FALSE}
plotPowerCont(outComputePower = pwr,  # 'computePower' output
              legendText = paste0("Cohort p = ", c(0.01, 0.02, 0.03, 0.05)))
```


